---
title: "Accessing Census' ACS Tribal Data via API"
format: html
editor: visual
---

## Background

This document describes API access to the Census Bureau's American Community Survey (ACS) statistical products for American Indian/Alaska Native/Native Hawaiian (AIANNH or "Tribal") geographies.

Census ACS products for the majority of AIAN geographies are limited to 5-year estimates due to the small population sizes of most AIANNH geographies. APIs are available for Detailed Tables, Subject Tables, and Data Profiles. (NOTE: Comparison Profile Tables are also available via API, but not for AIANNH geographies.) Therefore, the code provided here is only relevant for accessing ACS 5-year data from the Detailed Tables, Subject Tables, and Data Profiles. The figure on page 23 of [this Census presentation](https://www2.census.gov/training-workshops/courses/intro-census-data-api/api-course-module-1-presentation.pdf) summarizes the differences in these products.[^1]

[^1]: https://www2.census.gov/training-workshops/courses/intro-census-data-api/api-course-module-1-presentation.pdf

ACS data reported for Tribal geographies includes non-Tribal members; the Census Bureau warns that "data users...need to understand that many American Indian and Alaska Native \[AIAN\] areas have substantial populations that are neither American Indian nor Alaska Native."[^2] Tools summarizing data from Tribal geographies, such as the Census Bureau's My Tribal Area site, "provide characteristics for the entire population living in the AIAN area (regardless of their race or whether people identify as American Indian or Alaska Native)."[^3]

[^2]: <https://www.census.gov/content/dam/Census/library/publications/2021/acs/acs_aian_handbook_2021_ch02.pdf>

[^3]: <https://www.census.gov/content/dam/Census/library/publications/2021/acs/acs_aian_handbook_2021_ch05.pdf>

For this reason, whenever possible, data at the Tribal geography level should be subset by race/ethnicity for purposes related to describing Tribal populations. Not all ACS topics in the Census Bureau's data.census.gov are disaggregated by race/ethnicity.[^4] Examples in this document will assist users in finding Census ACS data for Tribal geographies disaggregated by race/ethnicity. Census reports racial categories as "race alone," when only one race is chosen, e.g., 'AIAN alone" or AIAN-A, and "race alone or in combination," when a one or more race is chosen, e.g. "AIAN alone or in combination" or AIAN-AoIC.

[^4]: <https://www.census.gov/content/dam/Census/library/publications/2021/acs/acs_aian_handbook_2021_ch02.pdf>

## Constructing API Calls from Census

### First get an API key

Census does not require users to have an API key, but there are limits on the number of requests that can come from a single IP address without a key. It's quick and easy to request a key: <https://api.census.gov/data/key_signup.html>.

```{r 1}
#| echo: false
# Assign your API key to a variable to call later
census_api_key="9b678913174c9e9d7825c60981d91a896e1dbdb6"
```

### Pull a table using the API

API calls can be constructed to either pull an entire table, analogous to downloading a table from data.census.gov, or to pull specific variables. We will first describe how to pull the table **B01001: Sex by Age** from the AIAN Detailed Tables. You can view this table in data.census.gov [here](https://data.census.gov/table/ACSDT5YAIAN2021.B01001?t=-02&g=010XX00US$2500000&d=ACS+5-Year+Estimates+American+Indian+and+Alaska+Native+Detailed+Tables). This version of the table includes all AIANNH geographies and data on "All available races alone or in combination". When you view the table in data.census.gov, you will see an "API" button in the header above the table, which provides the API endpoint. Copying and pasting this endpoint into a web browser will show you the data the table contains in (non-standard) JSON format (add *&outputFormat=csv* to the end of the URL to get the data in comma delimited format). When you construct an API call, you construct a similar endpoint and pull the data into your R environment and into a tidy and human readable data frame.

In this example, if you click on the API button when you open the table in data.census.gov, you'll get this URL

[`https://api.census.gov/data/2021/acs/acs5/aian?get=group(B01001)&POPGROUP=pseudo(-02)&ucgid=pseudo(0100000US$2500000)`](https://api.census.gov/data/2021/acs/acs5/aian?get=group(B01001)&POPGROUP=pseudo(-02)&ucgid=pseudo(0100000US$2500000))

The URL for the API call can be broken down into five parts:

1.  The base URL, in this example: *api.census.gov/data/2021/acs/acs5/aian*;

2.  The table id, in this example: *group(B01001)*;

3.  The population group iteration code, in this example: *POPGROUP=pseudo(-02)*;

4.  The geography code, in this example: *ucgid=pseudo(0100000US\$2500000)*; and

5.  The API key, appended at the end of the URL with *&key={API_KEY here}*.

Using the Census API requires you to know the table that you want to download, as well as understand Census codes for population groups and geographies.

1.  **Population group iteration codes**

These codes are used to subset data by racial category, but they are not the same as Census' race/ethnicity codes. They allow users to specify "X race alone," "X race alone or in combination," as well as specific tribal groups, e.g., "Eastern tribes, alone" (285) or "Yup'ik, alone" (190). The full list of codes can be found here: [iteration codes](https://www2.census.gov/programs-surveys/decennial/datasets/summary-file-2/attachment.pdf). In addition to these specific codes, common *sets* of codes can be pulled this the following:

```         
pseudo(-00) == “All available races” – this includes AIAN-A, AIAN-AoIC, and all specific tribal groups alone or in combination.

pseudo(-01) == “All available races alone” – this includes AIAN-A, but not AIAN-AoIC.

pseudo(-02) == “All available races alone or in combination” – this includes AIAN-AoIC, but not AIAN-A.

For pulling both AIAN-A and AIAN-AoIC without the specific tribal groupings, use pseudo(-01, -02). This is equivalent to pulling POPGROUPs **006** and **009** (see iteration codes).
```

2.  **Geographic codes**

Geographic codes are used to select the Census geographies for which you would like data summaries. To extract data for a specific geography, you need to know the associated GEO_ID.

-   The format for the call is “ucgid = {GEO_ID}," for example, for the Navajo Nation, your would specify “ucgid=2500000US2430”

-   To pull multiple geographies, use the “pseudo()” code convention:

    -   All AIANNH: pseudo(0100000US\$2500000).

    -   Only Alaska Native Regional Corporations (ANRCs): pseudo(0400000US02\$2300000)

We can put this all together into some code examples. These examples use the r packages `httr` and `jsonlite` to help construct the endpoints and then pull the data, `rvest` for reading in html content, and `tidyverse` for general tidy coding. If necessary, run `install.packages(c('tidyverse', 'httr', 'jsonlite', 'rvest'))` to install these packages before moving on.

```{r 2}
#| warning: false
#| message: false
# Call necessary packages
library(tidyverse)
library(httr)
library(jsonlite)
library(rvest) 
```

#### Example 1: Pull the B01001 table from the American Indian Alaska Native Tables (AIANT) for all tribal geographies and for the race/ethnicities "All available races alone or in combination"

```{r 3}
# Pull the data from the specific endpoint
# Note: API key is not used in this example
response <- GET('https://api.census.gov/data/2021/acs/acs5/aian?get=group(B01001)&POPGROUP=pseudo(-02)&ucgid=pseudo(0100000US$2500000)')
data <- fromJSON(content(response, "text"))

# Convert to data frame
df <- as.data.frame(data, stringsAsFactors = FALSE)
names(df) <- df[1, ]
df <- df[-1, ]

# Move identifying variables to the top of the df
df <- relocate(df, 'GEO_ID', 'NAME', 'POPGROUP', 'POPGROUP_LABEL')

```

```{r 3a}
#| echo: false
#| message: false
#| warning: false

# Show output of the df created above
library(DT)

datatable(
  head(df, 50),
  extensions = c("FixedColumns"),
  options = list(
    scrollX = TRUE,
    scrollY = "320px",
    paging  = FALSE,        # no pagination; we rely on vertical scroll
    autoWidth = TRUE
  ),
  class = "cell-border stripe hover"
)

```

It's hard to understand what exactly is being shown in this table but before diving into understanding the variables, let's look another simplified example of pulling an entire table from the AIANT.

#### Example 2: Pull the B01001 table from the American Indian Alaska Native Tables (AIANT) for all tribal geographies and for both the race/ethnicities "All available races alone" and "All available races alone or in combination".

```{r 4}
# Pull the data from the specific endpoint
# Note: API key is not used in this example
response <- GET('https://api.census.gov/data/2021/acs/acs5/aian?get=group(B01001)&POPGROUP=pseudo(-01,-02)&ucgid=pseudo(0100000US$2500000)')
# Note the addition of '-01' to the POPGROUP pseudo above
data <- fromJSON(content(response, "text"))

# Convert to data frame
df <- as.data.frame(data, stringsAsFactors = FALSE)
names(df) <- df[1, ]
df <- df[-1, ]

# Move identifying variables to the top of the df
df <- relocate(df, 'GEO_ID', 'NAME', 'POPGROUP', 'POPGROUP_LABEL')

```

```{r 4a}
#| echo: false
#| message: false
#| warning: false

# Show output of the df created above
library(DT)

datatable(
  head(df, 50),
  extensions = c("FixedColumns"),
  options = list(
    scrollX = TRUE,
    scrollY = "320px",
    paging  = FALSE,        # no pagination; we rely on vertical scroll
    autoWidth = TRUE
  ),
  class = "cell-border stripe hover"
)
```

If you scroll down in the dataframe, you can see that there are two different POPGROUPs included.

Let's look at one more example using this endpoint where we pull the most expansive set of race/ethnicity data for a single geography.

#### Example 3: Pull the B01001 table from the American Indian Alaska Native Tables (AIANT) for all race/ethnicities for a single AIANAH geography (Navajo Nation).

```{r 5}
# Pull the data from the specific endpoint
# Note: API key is not used in this example
response <- GET('https://api.census.gov/data/2021/acs/acs5/aian?get=group(B01001)&POPGROUP=pseudo(-00)&ucgid=2500000US2430')
# Note the use of '-00' in the POPGROUP pseudo above
data <- fromJSON(content(response, "text"))

# Convert to data frame
df <- as.data.frame(data, stringsAsFactors = FALSE)
names(df) <- df[1, ]
df <- df[-1, ]

# Move identifying variables to the top of the df
df <- relocate(df, 'GEO_ID', 'NAME', 'POPGROUP', 'POPGROUP_LABEL')
```

```{r 5a}
#| echo: false
#| message: false
#| warning: false

# Show output of the df created above
library(DT)

datatable(
  head(df, 50),
  extensions = c("FixedColumns"),
  options = list(
    scrollX = TRUE,
    scrollY = "320px",
    paging  = FALSE,        # no pagination; we rely on vertical scroll
    autoWidth = TRUE
  ),
  class = "cell-border stripe hover"
)
```

One thing to note about these examples is that when you pull an entire table, the dataframe that you end up is a transposition of the table as viewed in data.census.gov (that is, what appears as rows in the data.census.gov table are columns - variables - in the dataframe). This will be helpful to know for constructing API calls that pull a specific variable, covered in the next section.

### Pull specific variables from a table

If you look at the variable names of all the examples from the previous section, you'll notice that they all follow a similar pattern. First there are the identifying variables (GEO_ID, NAME, POPGROUP, POPGROUP_LABEL) - these typically default to the end of the dataframe and moving them to the front is the final step in the code, since that makes the data easier to read and use. After that, the variables have names that seem nonintuitive such as "B01001_001E". All variables in the Census ACS APIs geneally follow the same naming structure:[^5]

[^5]: See <https://www.census.gov/content/dam/Census/library/publications/2020/acs/acs_api_handbook_2020.pdf>, p7

**`{table ID}_{line number}{estimate type}`**

However, there are nuances and variations within this general format. Reading the section on "Understanding Variable Names" in the Census API Handbook (starting on page 7) is recommended.

"Estimate type" values are:

-   “E” refers to a numeric representation of the ACS estimate.

-   “M” refers to a numeric representation of the margin of error.

-   “PE” refers to an estimate representing a percent of the total.

-   “PM” refers to the margin of error for a percentage.

-   Anything with an "A" at the end refers to a special annotation of that estimate type (e.g. EA means an estimate annotation).

In the examples above:

-   B01001 = table ID (Sex by Age)

-   001 through 049 = the table line numbers, where 001 corresponds with the first line (Total) and 049 corresponds with the last line (85 years and over)

-   E = estimate, M = margin of error (MoE), EA and MA are annotations for those

This means that if you know the table that contains a specific variable you are interested in, you can construct the API call to pull just that variable, and you can determine the variable name based on the information that you know about it (table ID, row number, estimate type).

In the following examples, we will use a more dynamic form for constructing the API endpoint and eventually create a generalized function to construct the endpoint.

#### Example 4: Pull the top line of the B01001 table from the AIANT for all race/ethnicities and for all Tribal geographies

In this example, we are interested in data similar to what we were looking at in Example 3, but we just need the total population estimates for all of the race/ethnicities (including Tribal groupings), not the breakouts by sex or age. However, we also want this data for all Tribal geographies.

Since we are interested in the top line, we know this is row number 001. The table is the same as the previous examples - B01001. We are interested in the estimate (E) and let's say we also want MoE (M). Putting this together, the variables we want to pull along with the identifying variables (GEO_ID, NAME, POPGROUP, POPGROUP_LABEL) are: B01001_001E and B01001_001M.

```{r 6}
# List the variables WITHOUT spaces between the variable names
variables <- 'GEO_ID,NAME,POPGROUP,POPGROUP_LABEL,B01001_001E,B01001_001M'
# In this example, we are using the AIANT so the most recent year is 2021
year <- '2021'

base_url <- paste0("https://api.census.gov/data/", year, "/acs/acs5/aian")
  query <- list(
    get = variables,
    "for" = "american indian area/alaska native area/hawaiian home land:*",
    key = census_api_key # defined in a previous step 
  )

  # Make request
  response <- GET(base_url, query = query)
  data <- fromJSON(content(response, "text"))

  # Convert to data frame
  df <- as.data.frame(data, stringsAsFactors = FALSE)
  names(df) <- df[1, ]
  df <- df[-1, ]
  
# Move identifying variables to the top of the df
df <- relocate(df, 'GEO_ID', 'NAME', 'POPGROUP', 'POPGROUP_LABEL')
```

```{r 6a}
#| echo: false
#| message: false
#| warning: false

# Show output of the df created above
library(DT)

datatable(
  head(df, 50),
  extensions = c("FixedColumns"),
  options = list(
    scrollX = TRUE,
    scrollY = "320px",
    paging  = FALSE,        # no pagination; we rely on vertical scroll
    autoWidth = TRUE
  ),
  class = "cell-border stripe hover"
)
```

To see the URL of the endpoint that gets generated from the code above, run this:

```{r 7}
# This step is optional - it is just to help the user see the final API endpoint that is constructed from the steps above. The url that is printed below can be pasted into a web browser and the user will see the data that is being pulled.  
final_url <- modify_url(base_url, query = query)
cat(final_url, "\n")
```

The URL that is printed after running the code chunk can be pasted into a web browser and you will see the same data. You can compare this URL to the ones used in the examples above and see how they have a similar structure.

### Pull specific variables without using a table

Finding a specific variable that you are interested in does not require going through a bunch of tables in data.census.gov and locating the variable you need and then figuring out its row number and constructing the variable name from that. Census provides a full list of variables for each set of tables/profiles that can be viewed at the links below:

[Data Profile Variables](https://api.census.gov/data/2023/acs/acs5/profile/variables.html) (<https://api.census.gov/data/2023/acs/acs5/profile/variables.html>)

[Subject Table Variables](https://api.census.gov/data/2023/acs/acs5/subject/variables.html) (<https://api.census.gov/data/2023/acs/acs5/subject/variables.html>)

[AIAN Table Variables](https://api.census.gov/data/2021/acs/acs5/aian/variables.html) ([https://api.census.gov/data/2021/acs/acs5/aian/variables.html](https://api.census.gov/data/2021/acs/acs5/aian/variables.htmlhttps://api.census.gov/data/2021/acs/acs5/aian/variables.html))

[Detailed Table Variables](https://api.census.gov/data/2023/acs/acs5/variables.html) (<https://api.census.gov/data/2023/acs/acs5/variables.html>)

For the Data Profiles and Subject Tables, we can read these tables into dataframes and, since we are interested in variables that specifically apply to American Indian and Alaska Native populations, filter for variables that contain the string 'American Indian and Alaska Native'.

#### Example 5: Pull variables from a Data Profile table for total population, AIAN alone population, and AIAN alone or in combination population for all Tribal geographies

Data Profiles provide summaries of ACS data for various social, economic, housing, and demographic characteristics for the United States, regions, divisions, states, counties, county subdivisions, places, metropolitan areas, American Indian and Alaska Native areas, and other geographic areas. ACS Data Profiles begin with the letters “DP.”[^6]

[^6]: <https://www.census.gov/content/dam/Census/library/publications/2020/acs/acs_general_handbook_2020.pdf>

The code chunk below pulls all of the variables that are in the Data Profile tables and then filters for the ones that include "American Indian and Alaska Native".

```{r 8}
#| warning: false
#| message: false

# Data Profile variables

# URL of the variables page 
url <- "https://api.census.gov/data/2023/acs/acs5/profile/variables.html"  

# Read the HTML content 
page <- read_html(url)

# Extract all tables from the page 
tables <- page |>
  html_elements("table")  

# Convert the table to a data frame 
variables_df <- tables[[1]] |>    
  html_table(fill = TRUE) |>   
  tibble::as_tibble(.name_repair = "unique") |>   
  slice(-1) |>   
  select(-last_col())  

# filter to only AIAN 
dpvar_aian <- variables_df |>   
  filter(str_detect(Label, "American Indian and Alaska Native"))

rm(url, page, tables)
```

```{r 8a}
#| echo: false
#| message: false
#| warning: false

# Show output of the df created above
library(DT)

datatable(
  head(dpvar_aian, 50),
  extensions = c("FixedColumns"),
  options = list(
    scrollX = TRUE,
    scrollY = "320px",
    paging  = FALSE,        # no pagination; we rely on vertical scroll
    autoWidth = TRUE
  ),
  class = "cell-border stripe hover"
)
```

Notice that all of the AIAN variables start with DP05 - that's because only the table "[DP05: ACS Demographic and Housing Estimates](https://data.census.gov/table/ACSDP5Y2023.DP05)" contains race data. We can adapt the previous example to pull specific variables from this table. For example, we can pull population estimates and their MoEs for each Tribal geography for:

-   Total population (row 0001)

-   One race - AIAN (row 0039)

-   Race alone or in combination with one or more other races - AIAN (row 0071)

Putting together what we know about variable name construction, these would have the variable names: DP05_0001E/M, DP05_0039E/M, and DP05_0071E/M. Also, the POPGROUP concept only applies to the AIANT or other special population tables, so the POPGROUP and POPGROUP_LABEL variables have been removed in the code below.

```{r 9}
# List the variables WITHOUT spaces between the variable names
variables <- 'GEO_ID,NAME,DP05_0001E,DP05_0001M,DP05_0039E,DP05_0039M,DP05_0071E,DP05_0071M'
# In this example, we are using the Data Profiles so the most recent year is 2023
year <- '2023'

# base url has a slightly different form than the AIANT so this is different from the past examples
base_url <- paste0("https://api.census.gov/data/", year, "/acs/acs5/profile")
  query <- list(
    get = variables,
    "for" = "american indian area/alaska native area/hawaiian home land:*",
    key = census_api_key # defined in a previous step 
  )

  # Make request
  response <- GET(base_url, query = query)
  data <- fromJSON(content(response, "text"))

  # Convert to data frame
  df <- as.data.frame(data, stringsAsFactors = FALSE)
  names(df) <- df[1, ]
  df <- df[-1, ]
  
# Move identifying variables to the top of the df
df <- relocate(df, 'GEO_ID', 'NAME')
```

```{r 9a}
#| echo: false
#| message: false
#| warning: false

# Show output of the df created above
library(DT)

datatable(
  head(df, 50),
  extensions = c("FixedColumns"),
  options = list(
    scrollX = TRUE,
    scrollY = "320px",
    paging  = FALSE,        # no pagination; we rely on vertical scroll
    autoWidth = TRUE
  ),
  class = "cell-border stripe hover"
)
```

#### Example 6: Pull variables from a Subject Table table for total population and AIAN populations for all Tribal geographies

Subject Tables include ACS data organized by subject area, providing an overview of the information that analysts most often receive requests for from data users. ACS Subject Tables begin with the letter “S.”[^7]

[^7]: <https://www.census.gov/content/dam/Census/library/publications/2020/acs/acs_general_handbook_2020.pdf>

There are a lot of Subject Tables and corresponding variables, and the variables also have a slightly different structure to their names than with the other table types. Many (all?) of these tables are cross-tabulations, where there are multiple columns per geography. For example, if you look at [Table S1903](https://data.census.gov/table?q=S1903&g=2500000US2430&d=ACS+5-Year+Estimates+Subject+Tables) (in this example, for the Navajo Nation), you can see there are three column headings for which estimates and MoEs are presented: Number, Percent Distribution, and Median Income. These column headers translate to the variable names as a component that comes after the Table ID and begins with "C" and is followed by a two-digit, sequential number starting with 01. Specifically, the Subject Table variable names take the format: **`{table ID}_{column header}_{line number}{estimate type}`**. In this example, if we wanted to pull a specific row for Median Income (the third column header), the variable name would take the form S1903_C03\_(row number of variable)(E/M).

Because there are so many variables within the Subject Tables, and the added layer of multiple columns, the best approach to using these may be to examine the list of variables and searching in the Concept field for a specific subject matter, and seeing if there is a racial breakout for that subject matter. For example, the code below will produce two dataframes: one with all of the Subject Table variables ("subvar_all") and one with just the Subject Table variables that contain the string "American Indian and Alaska Native" in the Label field ("subvar_aian").

```{r 10}
#| warning: false
#| message: false

# Subject Table variables

# URL of the variables page 
url <- "https://api.census.gov/data/2023/acs/acs5/subject/variables.html"  

# Read the HTML content 
page <- read_html(url)

# Extract all tables from the page 
tables <- page |>
  html_elements("table")  

# Convert the table to a data frame 
subvar_all <- tables[[1]] |>    
  html_table(fill = TRUE) |>   
  tibble::as_tibble(.name_repair = "unique") |>   
  slice(-1) |>   
  select(-last_col())  

# filter to only AIAN 
subvar_aian <- subvar_all |>   
  filter(str_detect(Label, "American Indian and Alaska Native"))

rm(url, page, tables)
```

```{r 10a}
#| echo: false
#| message: false
#| warning: false

# Show output of the df created above
library(DT)

datatable(
  head(subvar_aian, 50),
  extensions = c("FixedColumns"),
  options = list(
    scrollX = TRUE,
    scrollY = "320px",
    paging  = FALSE,        # no pagination; we rely on vertical scroll
    autoWidth = TRUE
  ),
  class = "cell-border stripe hover"
)
```

Let's say we are interested in measures of median income in Tribal geographies. Searching the Concept field (using the built-in filter function in R Studio when viewing the subvar_all dataframe), we can see that there is a table (S1903) with a column header on Median income (C03) where the first row (*001E) is "Estimate!!Median income (dollars)!!HOUSEHOLD INCOME BY RACE AND HISPANIC OR LATINO ORIGIN OF HOUSEHOLDER!!Households"* and where another row within that table group (S1903_C03_004E) is "*Estimate!!Median income (dollars)!!HOUSEHOLD INCOME BY RACE AND HISPANIC OR LATINO ORIGIN OF HOUSEHOLDER!!Households!!One race--!!American Indian and Alaska Native*". We could pull both of those to view not only the median income of the households that identify as AIAN alone but also how that compares to the median income of all households in that Tribal geography, in case there is a significant proportion of non-Tribal members living within a geography.

```{r 11}
# List the variables WITHOUT spaces between the variable names
variables <- 'GEO_ID,NAME,S1903_C03_001E,S1903_C03_004E'
# In this example, we are using the Subject Tables so the most recent year is 2023
year <- '2023'

# base url has a slightly different form than the AIANT and Data Profiles so this is different from the past examples
base_url <- paste0("https://api.census.gov/data/", year, "/acs/acs5/subject")
  query <- list(
    get = variables,
    "for" = "american indian area/alaska native area/hawaiian home land:*",
    key = census_api_key # defined in a previous step 
  )

  # Make request
  response <- GET(base_url, query = query)
  data <- fromJSON(content(response, "text"))

  # Convert to data frame
  df <- as.data.frame(data, stringsAsFactors = FALSE)
  names(df) <- df[1, ]
  df <- df[-1, ]
  
# Move identifying variables to the top of the df
df <- relocate(df, 'GEO_ID', 'NAME')
```

```{r 11a}
#| echo: false
#| message: false
#| warning: false

# Show output of the df created above
library(DT)

datatable(
  head(df, 50),
  extensions = c("FixedColumns"),
  options = list(
    scrollX = TRUE,
    scrollY = "320px",
    paging  = FALSE,        # no pagination; we rely on vertical scroll
    autoWidth = TRUE
  ),
  class = "cell-border stripe hover"
)
```

When we examine the resulting data, we see there are instances where the value is "-666666666". It's not obvious what this means (something satanic?) but if you inspect the same base table in data.census.gov (in this example, [S1903](https://data.census.gov/table/ACSST5Y2023.S1903?q=S1903&g=010XX00US$2500000&d=ACS+5-Year+Estimates+Subject+Tables)) for Tribal geographies, you will see dashes for some of those geographies which indicates "The estimate could not be computed because there were an insufficient number of sample observations. For a ratio of medians estimate, one or both of the median estimates falls in the lowest interval or highest interval of an open-ended distribution. For a 5-year median estimate, the margin of error associated with a median was larger than the median itself." Essentially, it means those data are not available.

#### Example 7: View AIAN Table variables

The AIAN Tables are released as part of the Selected Population Tables, which are released every five years and provide the greatest level of statistical detail for specific population groups. The AIAN Tables show more detailed Tribal categories than the Selected Population Tables.[^8] However, because it is only released every five years, the most recent data available are from 2021.

[^8]: <https://www.census.gov/programs-surveys/acs/data/race-aian.html>

For the AIAN Tables, *all* variables can be broken out by various types of American Indian and Alaska Native identifiers or Tribal groups (the POPGROUP concept discussed in examples 1 through 4) so any of the variables available in the AIANT endpoint can be augmented using POPGROUP to view specific AIAN/Tribal grouping estimates. The code chunk below is for creating a dataframe of all variables in these tables; for example on pulling variables from this endpoint, refer to examples 1 through 4.

```{r 12}
#| warning: false
#| message: false

# AIAN Table variables

url <- "https://api.census.gov/data/2021/acs/acs5/aian/variables.html"

# Read the HTML content 
page <- read_html(url)

# Extract all tables from the page 
tables <- page |>
  html_elements("table")  

# Convert the table to a data frame 
aiantvar <- tables[[1]] |>    
  html_table(fill = TRUE) |>   
  tibble::as_tibble(.name_repair = "unique") |>   
  slice(-1) |>   
  select(-last_col())

rm(page, tables)
```

```{r 12a}
#| echo: false
#| message: false
#| warning: false

# Show output of the df created above
library(DT)

datatable(
  head(aiantvar, 50),
  extensions = c("FixedColumns"),
  options = list(
    scrollX = TRUE,
    scrollY = "320px",
    paging  = FALSE,        # no pagination; we rely on vertical scroll
    autoWidth = TRUE
  ),
  class = "cell-border stripe hover"
)
```

#### Example 8: Pull data on Median Age of AIAN Alone populations from all Tribal geographies from a Detailed Table

Detailed Tables provide access to the most comprehensive ACS data tables on all topics and geographic areas. They begin with the letters “B” for base tables and “C” for collapsed tables. The “collapsed” tables cover the same topics as the base table, but with fewer categories.[^9] In addition, the table ID sometimes ends with the letter A through I - these indicate tables that are broken out for specific race/ethnicity categories, where "C" corresponds with the race/ethnicity category "American Indian and Alaska Native Alone". If using these tables, it is a best practice to first compare the estimates of American Indian and Alaska Native Alone with American Indian and Alaska Native Alone or in Combination for that geography (using table DP05) in order to determine whether the geography has a significant portion of potential Tribal members who are identifying as multiracial.

[^9]: <https://www.census.gov/content/dam/Census/library/publications/2020/acs/acs_general_handbook_2020.pdf>

Otherwise, the format of these variables takes the same general form that has been used in most of the other tables: {TableID}\_{Row Number}{Estimate Type}. Similar to what was done for the Subject Tables, a user could approach these tables by viewing and filtering dataframes of the variables to find those that contain a topic of interest, and seeing if there is a racial breakout. The code below will produce two dataframes: one with all of the Detailed Table variables ("dtvar_all") and one with just the Detailed Table variables that contain the string "American Indian and Alaska Native" in the Label or Concept fields ("dtvar_aian").

```{r 13}
#| warning: false
#| message: false

# Detailed Table variables

url <- "https://api.census.gov/data/2023/acs/acs5/variables.html"

# Read the HTML content 
page <- read_html(url)

# Extract all tables from the page 
tables <- page |>
  html_elements("table")  

# Convert the first table to a data frame 
dtvar_all <- tables[[1]] |>    
  html_table(fill = TRUE) |>   
  tibble::as_tibble(.name_repair = "unique") |>   
  slice(-1) |>   
  select(-last_col())

# filter to only AIAN 
# The Detailed Tables differ from the Subject Tables and Data Profiles in that race can appear in both the Label field and the Concept field
dtvar_aian <- dtvar_all |>   
  filter(str_detect(Label, "American Indian and Alaska Native") | str_detect(Concept, "American Indian and Alaska Native"))

rm(page, tables, url)
```

```{r 13a}
#| echo: false
#| message: false
#| warning: false

# Show output of the df created above
library(DT)

datatable(
  head(dtvar_aian, 50),
  extensions = c("FixedColumns"),
  options = list(
    scrollX = TRUE,
    scrollY = "320px",
    paging  = FALSE,        # no pagination; we rely on vertical scroll
    autoWidth = TRUE
  ),
  class = "cell-border stripe hover"
)
```

We can review the list of variables available for AIAN populations and pick one, like B01002C_001E (Median Age by Sex (American Indian and Alaska Native Alone)), and pull the data for all Tribal geographies:

```{r 14}
# List the variables WITHOUT spaces between the variable names
variables <- 'GEO_ID,NAME,B01002C_001E'
# In this example, we are using the Detailed Tables so the most recent year is 2023
year <- '2023'

# base url has a slightly different form than the AIANT and Data Profiles so this is different from the past examples
base_url <- paste0("https://api.census.gov/data/", year, "/acs/acs5")
  query <- list(
    get = variables,
    "for" = "american indian area/alaska native area/hawaiian home land:*",
    key = census_api_key # defined in a previous step 
  )

  # Make request
  response <- GET(base_url, query = query)
  data <- fromJSON(content(response, "text"))

  # Convert to data frame
  df <- as.data.frame(data, stringsAsFactors = FALSE)
  names(df) <- df[1, ]
  df <- df[-1, ]
  
# Move identifying variables to the top of the df
df <- relocate(df, 'GEO_ID', 'NAME')
```

```{r 14a}
#| echo: false
#| message: false
#| warning: false

# Show output of the df created above
library(DT)

datatable(
  head(df, 50),
  extensions = c("FixedColumns"),
  options = list(
    scrollX = TRUE,
    scrollY = "320px",
    paging  = FALSE,        # no pagination; we rely on vertical scroll
    autoWidth = TRUE
  ),
  class = "cell-border stripe hover"
)
```

While the Census documentation indicates that the Detailed Tables cover "All areas down to the block group level"[^10], this does not appear to be true as there are variables that do not return values for AIANNH geographies but do for other geographies, such as states. For example, compare the results in the code chunk below, which pulls estimates of the AIAN population who work from home, by commenting/uncommenting the "for" lines in the query of the base_url:

[^10]: Understanding and Using American Community Survey Data: What All Data Users Need to Know, p21

```{r 15}
# List the variables WITHOUT spaces between the variable names
variables <- 'GEO_ID,NAME,B08505C_007E'
# In this example, we are using the Detailed Tables so the most recent year is 2023
year <- '2023'

# base url has a slightly different form than the AIANT and Data Profiles so this is different from the past examples
base_url <- paste0("https://api.census.gov/data/", year, "/acs/acs5")
  query <- list(
    get = variables,
    #"for" = "american indian area/alaska native area/hawaiian home land:*",
    "for" = "state:*",
    key = census_api_key # defined in a previous step 
  )

  # Make request
  response <- GET(base_url, query = query)
  data <- fromJSON(content(response, "text"))

  # Convert to data frame
  df <- as.data.frame(data, stringsAsFactors = FALSE)
  names(df) <- df[1, ]
  df <- df[-1, ]
  
# Move identifying variables to the top of the df
df <- relocate(df, 'GEO_ID', 'NAME')
```

```{r 15a}
#| echo: false
#| message: false
#| warning: false

# Show output of the df created above
library(DT)

datatable(
  head(df, 50),
  extensions = c("FixedColumns"),
  options = list(
    scrollX = TRUE,
    scrollY = "320px",
    paging  = FALSE,        # no pagination; we rely on vertical scroll
    autoWidth = TRUE
  ),
  class = "cell-border stripe hover"
)
```

It's a good idea to use the advanced search tool for data.census.gov (<https://data.census.gov/advanced>) to get a sense for what kinds of data are available for the topic you are interested in, because there may be multiple tables that address that topic and those tables may be available for different geographies. For example, though data on AIAN populations working from home were not available for Tribal geographies through the Detailed Tables, using the advanced search tool would point you to a Subject Table (S0802) that has similar types of data for Tribal geographies that could potentially be used instead.
