---
title: "CensusTribalWorkflow"
format: html
editor: visual
---

```{r}
#| echo: false
# Assign your API key to a variable to call later
# Will run with census_api_key set to '' but there can be limits on the number of requests
#census_api_key='' #note: uncomment before publishing
census_api_key='9b678913174c9e9d7825c60981d91a896e1dbdb6' #note: this is Ann's key, remove before publishing
```

```{r}
#| warning: false
#| message: false
# Call necessary packages
library(tidyverse)
library(httr)
library(jsonlite)
library(rvest) 
```

## Step 1: Pull Decennial data at the tribal geography level for total population, AIAN-alone population, and AIAN-AoIC population

```{r}
api_url <- paste0('https://api.census.gov/data/2020/dec/dhc?get=group(P8)&ucgid=pseudo(0100000US$2500000)', '&descriptive=true', '&key=', census_api_key)
response <- GET(api_url)
data <- fromJSON(content(response, "text"))

# Convert to data frame
df <- as.data.frame(data, stringsAsFactors = FALSE)
names(df) <- df[1, ]
df <- df[-1, ]

```

```{r}
# Figure out which variables we want to select. We need a measure of total population and then any variables that are associated with populations identifying as AIAN.
vars <- df[1, ] |>
  pivot_longer(
    cols = everything(),
    names_to = 'var_name',
    values_to = 'var_desc'
  )

aian_vars <- vars |>
  filter(str_detect(var_desc, 'American Indian'))|> # filter to variables with 'American Indian' in the description
  filter(grepl("N$", var_name)) # filter to variables where the variable ends with N (instead of NA which represents an annotation)

select_vars <- aian_vars$var_name
select_vars <- c('GEO_ID', 'NAME', 'ucgid', 'P8_001N', select_vars) # Add the variable for total population and geog identifying vars


# Select the variables identified above and sum all the "two or more races" variables to create one AIAN_AOIC variable
df_sum <- df |>
  select(any_of(select_vars)) |>
  filter(GEO_ID != 'Geography') |> # drop first row that contains variable descriptions 
  mutate(across(P8_001N:last_col(), as.numeric)) |>
  rowwise() |>
  mutate(dec_pop_aian_ic = sum(c_across(P8_012N:last_col()), na.rm = TRUE)) # sum two or more races vars; should consider making this more robust - this depends on the df being structured in a certain way (having all the two or more races vars clustered together)

dec_pop <- df_sum |>
  select(GEO_ID, NAME, P8_001N, P8_005N, dec_pop_aian_ic) |>
  rename('dec_pop_total' = 'P8_001N',
         'dec_pop_aian_a' = 'P8_005N') |>
  mutate(dec_pct_aian = (dec_pop_aian_a + dec_pop_aian_ic) / dec_pop_total)
```

## Step 2: Pull the most recent ACS-5 data at the tribal geography level for total population, AIAN-alone population, and AIAN-AoIC population

```{r}
# Data Profile variables

# URL of the variables page 
url <- 'https://api.census.gov/data/2024/acs/acs5/profile/variables.html'

# Read the HTML content 
page <- read_html(url)

# Extract all tables from the page 
tables <- page |>
  html_elements("table")  

# Convert the table to a data frame 
variables_df <- tables[[1]] |>    
  html_table(fill = TRUE) |>   
  tibble::as_tibble(.name_repair = "unique") |>   
  slice(-1) |>   
  select(-last_col())

selected_vars <- variables_df |>
  filter(Label == 'Estimate!!RACE!!Total population' |
            Label == 'Estimate!!RACE!!Total population!!One race!!American Indian and Alaska Native' |
              Label == 'Estimate!!Race alone or in combination with one or more other races!!Total population!!American Indian and Alaska Native')

select_vars <- selected_vars$Name

rm(url, page, tables)
```

```{r}
#| warning: false
#| message: false

# List the variables WITHOUT spaces between the variable names
variables <- c('GEO_ID', 'NAME', select_vars) # Add geog identifying vars to the selected vars identified above
variables <- paste(variables, collapse = ',') # Collapse into a string with no spaces
  'GEO_ID,NAME,DP05_0001E,DP05_0001M,DP05_0039E,DP05_0039M,DP05_0085E,DP05_0085M'
# In this example, we are using the Data Profiles so the most recent year is 2024
year <- '2024'

# base url has a slightly different form than the AIANT so this is different from the past examples
base_url <- paste0('https://api.census.gov/data/', year, '/acs/acs5/profile')
  query <- list(
    get = variables,
    'for' = 'american indian area/alaska native area/hawaiian home land:*',
    descriptive = 'true',
    key = census_api_key # defined in a previous step 
  )

  # Make request
  response <- GET(base_url, query = query)
  data <- fromJSON(content(response, "text"))

  # Convert to data frame
  df <- as.data.frame(data, stringsAsFactors = FALSE)
  names(df) <- df[1, ]
  df <- df[-1, ]
  
# Move identifying variables to the top of the df
df <- relocate(df, 'GEO_ID', 'NAME')
```

```{r}
# Repeat the computations that were done with the decennial data for comparison
acs_pop <- df |>
  rename('acs_pop_total' = 'DP05_0033E',
         'acs_pop_aian_a' = 'DP05_0053E',
         'acs_pop_aian_aoic' = 'DP05_0085E') |>
  mutate(across(acs_pop_total:last_col(), as.numeric)) |>
  select(GEO_ID, NAME, acs_pop_total, acs_pop_aian_a, acs_pop_aian_aoic) |>
  filter(GEO_ID != 'Geography') |> # drop first row that contains variable descriptions 
  mutate(acs_pct_aian = (acs_pop_aian_aoic) / acs_pop_total)
  
```

## Compare total population count versus most recent estimate

```{r}
total_pop <- dec_pop |>
  select(GEO_ID, NAME, dec_pop_total) |>
  left_join(select(acs_pop, GEO_ID, acs_pop_total)) |>
  mutate(pct_chg = acs_pop_total/dec_pop_total -1)
```

## Compare percent AIAN in Decennial versus most recent ACS

```{r}
pct_aian <- dec_pop |>
  select(GEO_ID, NAME, dec_pct_aian) |>
  left_join(select(acs_pop, GEO_ID, acs_pct_aian)) |>
  mutate(dec_acs_comp = dec_pct_aian / acs_pct_aian)
```
