---
title: "Census tribal pop function use cases"
format: html
editor: visual
---

# Using Census tribal population functions

## Description

The purpose of this script is to provide examples and use cases for the functions in the census-tribal repo.

This script uses functions developed by the DOI Office of Policy Analysis to pull the most recent Decennial (2020) and 5-year American Community Survey (ACS, 2024) population data for American Indian/Alaska Native Census geographies that are associated with federally-recognized tribes. Aside from the total population, it also pulls the population of people identifying as American Indian or Alaska Native (AIAN) Alone ("A") and Alone or in Combination ("AoIC"), calculates the percentage of the population that selected AIAN as at least one of the races/ethnicities that they identify with ("pct_AIAN"), and calculates the ratio of individuals identifying as AIAN Alone to individuals identifying as multiracial AIAN (AIAN Alone or in Combination).

```{r}
# load libraries
# if necessary, install libraries but uncommenting the next line:
#install.packages(c("tidyverse", "here", "tidycensus"))
library(tidyverse)
library(here)
library(tidycensus)
#library(DT)
```

### Load functions and set Census API key

Census does not require users to have an API key, but there are limits on the number of requests that can come from a single IP address without a key. It's quick and easy to request a key: <https://api.census.gov/data/key_signup.html>.

```{r}
# Paste your key between the quotation marks below:
census_api_key('9b678913174c9e9d7825c60981d91a896e1dbdb6') #Ann's key, delete before publishing

source(here::here("src", "get_popn_data_DECENNIAL_fn.R"))
source(here::here("src", "get_popn_data_ACS_fn.R"))

```

### Pull Decennial data (2020) for all tribal geographies associated with federally recognized tribes

Variable definitions (all variables beginning with "dec" refer to data from the 2020 Decennial):

-   dec.NAME - The name of the geography

-   dec.total - The total population of the geography

-   dec.A - The population of the geography identifying as AIAN Alone

-   dec.AoIC - The population of the geography identifying as AIAN Alone or in Combination

-   dec.pct_AIAN - The percentage of the total population identifying as AIAN Alone or in Combination

-   dec.A_AoIC - The ratio of individuals identifying as AIAN Alone to individuals identifying as AIAN Alone or in Combination (a measure of how much of the AIAN population identifies as multiracial)

```{r}
dec <- get_popn_data() |>
  pivot_wider(id_cols = c("GEOID", "NAME"), names_from = variable) |>
  mutate(AoIC = A + IC) |>
  select(GEOID, NAME, total, A, AoIC) |>
  mutate(dec.pct_AIAN = AoIC / total) |>
  mutate(dec.A_AoIC = A / AoIC) |>
  rename(dec.total = total,
         dec.A = A,
         dec.AoIC = AoIC,
         dec.NAME = NAME)
```

### Pull the most recent ACS data (2024) for all tribal geographies associated with federally recognized tribes

Variable definitions (all variables beginning with "acs" refer to data from the most recent 5-year ACS):

-   acs.NAME - The name of the geography

-   acs.total - The total population of the geography

-   acs.A - The population of the geography identifying as AIAN Alone

-   acs.AoIC - The population of the geography identifying as AIAN Alone or in Combination

-   acs.pct_AIAN - The percentage of the total population identifying as AIAN Alone or in Combination

-   acs.A_AoIC - The ratio of individuals identifying as AIAN Alone to individuals identifying as AIAN Alone or in Combination (a measure of how much of the AIAN population identifies as multiracial)

```{r}
acs24 <- get_acs_popn_data(year = 2024) |> # need to specify the year when calling the function
  pivot_wider(id_cols = c("GEOID", "NAME"), names_from = variable, values_from = estimate) |>
  select(GEOID, NAME, total, A, AoIC) |>
  mutate(acs.pct_AIAN = AoIC / total) |>
  mutate(acs.A_AoIC = A / AoIC) |>
  rename(acs.total = total,
         acs.A = A,
         acs.AoIC = AoIC,
         acs.NAME = NAME)
  
```

### Compile Decennial and ACS data for each tribal geography and export to .csv for use

```{r}
dec_acs24 <- dec |>
  full_join(acs24, by = 'GEOID')

# EMILY: where do we want the output .csv file to go? Currently it's being directed to the 'data' folder
#write.csv(dec_acs24, here('data', 'census-tribal_Decennail_ACS24_Population.csv'))

```
